---
title: "jared_is_back2"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, cache=TRUE, fig.align = "center")
options(digits=5, pillar.sigfig = 7)
library(tidyverse)
library(tidymodels)
library(dplyr)
library(gganimate)
library(gapminder)
```

## Import Longitude-Latitude Dataset

```{r}
lats_long <- read.csv("https://raw.githubusercontent.com/albertyw/avenews/master/old/data/average-latitude-longitude-countries.csv")
```

```{r}
lats_long <- lats_long %>%
  rename(location = Country)
```

## Import John Hopkins Covid Dataset from 12/5/2021 named `covid`

```{r}
covid <- read.csv("https://raw.githubusercontent.com/ST47S-CompStats-Fall2021/GroupJ-COVID/main/data/covid-12-5-2021?token=AST7DELZDJCUNN4DQ752EDLBXAASG")

# change date to date variable
covid <- covid %>%
  mutate(date = as.Date(date))

# add longitude-latitude to covid dataset
covid <- left_join(covid, lats_long, by = "location")
```

## Functions

#### `filter_continents` : filter covid dataset with `continent` variable for only sevon seperate continents 

```{r}
filter_continents <- function(covid_data) {
   covid_data <- covid_data %>%
     filter(continent == "Asia" |
           continent == "Africa" |
           continent == "Europe" |
           continent == "North America" |
           continent == "Oceania" |
           continent == "South America" |
           continent == "Antarctica"
           )
}
```

#### `only_countries` : filter covid dataset with `continent` variable for only sevon seperate continents 

```{r}
filter_world <- function(covid_data) {
  covid_data <- covid_data %>%
    filter(location != "World" |
           location != "Asia" |
           location != "Africa" |
           location != "Europe" |
           location != "North America" |
           location != "Oceania" |
           location != "South America" |
           location != "Antarctica"
           )
}
```

#### `replace_all_na` : replaces all `NA`'s in a dataframe with 0s

```{r}
replace_all_na <- function(covid_data) {
  covid_data %>%
    replace(is.na(.), 0)
}
```

## Creating Sub-datasets

#### `covid_numeric` : Only numeric variables (and iso_code, location, date, continent)

```{r}
#select only numeric variables (outside of iso_code, location, and date)
covid_numeric <- cbind(
  covid %>%
    select(iso_code, location, date, continent),
  covid %>%
    select_if(is.numeric)
)
```

### `covid_per` : Only variables that containing `per` (and iso_code, location, date, continent)

```{r}
covid_per <- cbind(
  covid %>%
    select(iso_code, location, date, continent, population),
  covid %>%
    select(contains("per"))
)
```

### `covid_pm` : Only variables that end in per_million (and iso_code, location, date, continent)

```{r}
covid_pm <- cbind(
  covid %>%
    select(iso_code, location, date, continent, population),
  covid %>%
    select(ends_with("per_million"))
)
```

### `covid_spm` : Only variables that end with smoothed_per_million (and iso_code, location, date)

```{r}
covid_spm <- cbind(
  covid %>%
    select(iso_code, location, date, continent),
  covid %>%
    select(ends_with("smoothed_per_million"))
)
```

#### `covid_spm_location` : (pivot wider)

```{r}
covid_spm_location <- covid_spm %>%
  pivot_wider(names_from = date, values_from = !c(iso_code, date, location, continent)) %>%
  replace_all_na()
```

## EDA - Visualizations

#### New cases Smooth Per Million Over Time By Continent
```{r}
covid_spm %>% head()
```


```{r}
covid_spm %>%
  filter_continents %>%
  ggplot(aes(x = date, y = new_cases_smoothed_per_million, color=location)) +
  geom_line(aes(color = location), show.legend = FALSE) +
  facet_wrap(~continent) +
  scale_colour_manual(values = country_colors)

```

```{r}
covid_spm %>%
  filter_continents %>%
  ggplot(aes(x = date, y = new_deaths_smoothed_per_million, color=location)) +
  geom_line(aes(color = location), show.legend = FALSE) +
  facet_wrap(~continent) +
  scale_colour_manual(values = country_colors)

```

```{r}
covid_spm %>%
  filter_continents %>%
  ggplot(aes(x = date, y = new_vaccinations_smoothed_per_million, color=location)) +
  geom_line(aes(color = location), show.legend = FALSE) +
  facet_wrap(~continent) +
  scale_colour_manual(values = country_colors)

```


## Misc/Experiments


#### gganimate example

```{r}
covid_spm1 <- covid_spm %>%
  replace_all_na() %>%
  filter_world()

covid_pm1 <- covid_pm %>%
  replace_all_na() %>%
  filter_world() %>%
  filter_continents()
```

```{r}
ggplot(covid_pm1, aes(total_cases_per_million, total_deaths_per_million, color = location, size = population)) +
  #geom_point(alpha = 0.7, show.legend = FALSE) +
  geom_smooth(size=0.5, se=FALSE, show.legend=FALSE) +
  facet_wrap(~continent) +
  scale_color_manual(values = country_colors)

```

```{r, eval=FALSE}
cases_death_animate <- ggplot(covid_pm1, aes(total_cases_per_million, total_deaths_per_million, color = location, size = population)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_color_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  facet_wrap(~continent) +
  labs(title = 'Date: {frame_time}', x = 'Total Cases per Million', y = 'Total Deaths per Million') +
  transition_time(date) +
  ease_aes('linear') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

animate(cases_death_animate, duration = 10)
```




#### gganimate experiment

##### cases per million
```{r}
world <- map_data("world")
```

```{r}
map <- ggplot() +
  geom_map(
    data = world, map = world,
    aes(long, lat, map_id = region),
    color = "black", fill = "lightgray", size = 0.01
  ) +
  geom_point(data = covid, aes(Longitude, Latitude, color=location, size=total_cases_per_million), show.legend = FALSE) +
  scale_color_manual(values = country_colors) +
  scale_size(range = c(0.1,5))
  
map
```


```{r, eval=FALSE}
map_anim <- ggplot() +
  geom_map(
    data = world, map = world,
    aes(long, lat, map_id = region),
    color = "black", fill = "lightgray", size = 0.01
  ) +
  geom_point(data = covid, aes(Longitude, Latitude, color=location, size=total_cases_per_million), show.legend = FALSE) +
  scale_color_manual(values = country_colors) +
  scale_size(range = c(0.1, 5)) +
  transition_time(date)

animate(map_anim, duration = 10)
```



#### cases total
```{r}
map <- ggplot() +
  geom_map(
    data = world, map = world,
    aes(long, lat, map_id = region),
    color = "black", fill = "lightgray", size = 0.01
  ) +
  geom_point(data = covid, aes(Longitude, Latitude, color=location, size=total_cases), show.legend = FALSE) +
  scale_color_manual(values = country_colors) +
  scale_size(range = c(1, 15))
  
map
```

```{r, eval=FALSE}
anim_map_total_cases <- ggplot() +
  geom_map(
    data = world, map = world,
    aes(long, lat, map_id = region),
    color = "black", fill = "lightgray", size = 0.01
  ) +
  geom_point(data = covid, aes(Longitude, Latitude, color=location, size=total_cases), show.legend = FALSE) +
  scale_color_manual(values = country_colors) +
  scale_size(range = c(1, 15)) +
  transition_time(date)

animate(anim_map_total_cases, duration=10)
```








