---
title: "jared_is_back"
author: "Jared Mejia"
date: "12/11/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, cache=TRUE, fig.align = "center")
options(digits=5, pillar.sigfig = 7)
library(tidyverse)
library(tidymodels)
library(dplyr)
library(gganimate)
```

# This is the return of Jared

## Catch-up

First, Jared needs to catchup with what Doc has done

### Import Longitude-Latitude Dataset

```{r}
lats_long <- read.csv("https://raw.githubusercontent.com/albertyw/avenews/master/old/data/average-latitude-longitude-countries.csv")
```

```{r}
lats_long <- lats_long %>%
  rename(location = Country)
```

```{r}
head(lats_long)
```

### Import Covid Dataset
```{r}
covid <- read.csv("https://raw.githubusercontent.com/ST47S-CompStats-Fall2021/GroupJ-COVID/main/data/covid-12-5-2021?token=AMFC5PUJYM3IRW4QZIBDMY3BXZCXO")

# converting wonky dates to variable formated as data
covid <- covid %>%
  mutate(date = as.Date(date))

head(covid)
```

Adding long-lat to the covid data by matching the `location` variable
```{r}
covid <- left_join(covid, lats_long, by="location")
head(covid)
```

Next we choose the data with valid continents:

```{r}
continents <- c("Asia", "Africa", "Europe", "North America", "Oceania", "South America", "Antarctica")

covid <- covid %>%
  filter(continent %in% continents)
```

```{r}
unique(covid$continent)
```

Doc's function to replace NA's with 0s
```{r}
replace_all_na <- function(covid_data) {
  covid_data %>%
    replace(is.na(.), 0)
}
```


Dataframe only including `smoothed_per_million` variables:
```{r}
covid_spm <- covid %>%
  select(iso_code, location, date, continent, ends_with("smoothed_per_million"))

head(covid_spm)
```

Plotting `smoothed_per_million` variables

```{r}
covid_spm %>%
  ggplot(aes(x = date, y = new_cases_smoothed_per_million)) +
  geom_line(aes(color = continent)) +
  facet_wrap(~continent)
```

Well, that kind of looks like a drawing made by a toddler. Let's try to plot animations over time instead:

```{r}
covid_spm <- cbind(
  covid_spm %>% select(date),
  covid_spm %>%
    select(!date) %>%
    mutate_all(~replace(., is.na(.), 0))
)

head(covid_spm)

```

```{r}

filter_continents <- function(covid_data) {
   covid_data <- covid_data %>%
     filter(continent == "Asia" |
           continent == "Africa" |
           continent == "Europe" |
           continent == "North America" |
           continent == "Oceania" |
           continent == "South America" |
           continent == "Antarctica"
           )
}

filter_world <- function(covid_data) {
  covid_data <- covid_data %>%
    filter(location != "World" |
           location != "Asia" |
           location != "Africa" |
           location != "Europe" |
           location != "North America" |
           location != "Oceania" |
           location != "South America" |
           location != "Antarctica"
           )
}

covid_spm <- cbind(
  covid %>%
    select(iso_code, location, date, continent),
  covid %>%
    select(ends_with("smoothed_per_million"))
)

covid_spm1 <- covid_spm %>%
  filter(!is.na(date)) %>%
  replace_all_na() %>%
  filter_world()
```


```{r}
library(gapminder)

new_cases_deaths_animate <- covid_spm %>%
  ggplot(aes(x=new_cases_smoothed_per_million, y=new_deaths_smoothed_per_million, color=location)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_color_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  facet_wrap(~continent) +
  labs(title = 'Date: {frame_time}', x = 'New Cases per Million (smoothed)', y = 'New Deaths per Million (smoothed)') +
  transition_time(date) +
  ease_aes('linear') +
  theme(axis.text.x = element_text(angle = 90, vjust=0.5))
```

```{r}
animate(new_cases_deaths_animate, duration = 10)
```

```{r}
cases_death_animate <- ggplot(covid_spm, aes(new_cases_per_million, new_deaths_per_million, color = location)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_color_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  facet_wrap(~continent) +
  labs(title = 'Date: {frame_time}', x = 'New Cases per Million (smoothed)', y = 'New Deaths per Million (smoothed)') +
  transition_time(date) +
  ease_aes('linear') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))


```


```{r}
animate(cases_death_animate, duration=10)
```


```{r}
covid_pm <- cbind(
  covid %>%
    select(iso_code, location, date, continent, population),
  covid %>%
    select(ends_with("per_million"))
)

covid_pm1 <- covid_pm %>%
  filter(!is.na(date)) %>%
  replace_all_na() %>%
  filter_world() %>%
  filter_continents()

cases_death_animate <- ggplot(covid_pm1, aes(total_cases_per_million, total_deaths_per_million, color = location)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_color_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  facet_wrap(~continent) +
  labs(title = 'Date: {frame_time}', x = 'total_deaths_per_million', y = 'new_cases_per_million') +
  transition_time(date) +
  ease_aes('linear') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
animate(cases_death_animate, duration = 10)
```

```{r}
animate(cases_death_animate, duration = 10)
```



```{r}
covid_spm <- cbind(
  covid %>%
    select(iso_code, location, date, continent),
  covid %>%
    select(ends_with("smoothed_per_million"))
)

covid_spm1 <- covid_spm %>%
  filter(!is.na(date)) %>%
  replace_all_na() %>%
  filter_world() %>%
  filter_continents()

new_cases_death_animate <- ggplot(covid_spm1, aes(new_cases_smoothed_per_million, new_deaths_smoothed_per_million, color = location)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_color_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  facet_wrap(~continent) +
  labs(title = 'Date: {frame_time}', x = 'New Cases per Million (smoothed)', y = 'New Deaths per Million (smoothed)') +
  transition_time(date) +
  ease_aes('linear') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

animate(new_cases_death_animate, duration = 10)
```

















