---
title: "COVID-EDA-Guy"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, cache=TRUE, fig.align = "center")
options(digits=5, pillar.sigfig = 7)
library(tidyverse)
library(tidymodels)
library(dplyr)
library(gganimate)
```

## Import John Hopkins Covid Dataset from 12/5/2021 named `covid`

```{r}
covid <- read.csv("https://raw.githubusercontent.com/ST47S-CompStats-Fall2021/GroupJ-COVID/main/data/covid-12-5-2021?token=AST7DELZDJCUNN4DQ752EDLBXAASG")

# change date to date variable
covid <- covid %>%
  mutate(date = as.Date(date))
```

## Creating Sub-datasets

#### `covid_numeric` : Only numeric variables (and iso_code, location, date, continent)

```{r}
#select only numeric variables (outside of iso_code, location, and date)
covid_numeric <- cbind(
  covid %>%
    select(iso_code, location, date, continent),
  covid %>%
    select_if(is.numeric)
)
```

### `covid_pm` : Only variables that end in per_million (and iso_code, location, date, continent)

```{r}
covid_pm <- cbind(
  covid %>%
    select(iso_code, location, date),
  covid %>%
    select(ends_with("per_million"))
)
```

### `covid_spm` : Only variables that end with smoothed_per_million (and iso_code, location, date)

```{r}
covid_spm <- cbind(
  covid %>%
    select(iso_code, location, date, continent),
  covid %>%
    select(ends_with("smoothed_per_million"))
)
```

#### `covid_spm_location` : (pivot wider)

```{r}
covid_spm_location <- covid_spm %>%
  pivot_wider(names_from = date, values_from = !c(iso_code, date, location, continent)) %>%
  replace_all_na()
```

```{r}
covid_spm_location_km <- covid_spm_location %>%
  select(is.numeric) %>%
  replace_all_na()

set.seed(47)
covid_spm_location_kclust <- covid_spm_location_km %>%
  kmeans(centers = 3)
```

##### Tuning k in k-means

```{r}
set.seed(47)
kmax <- 30
covid_kclusts <- 
  tibble(k = 1:kmax) %>%
  mutate(
    covid_kclust = map(k, ~kmeans(covid_spm_location_km, .x)),
    tidied = map(covid_kclust, tidy),
    glanced = map(covid_kclust, glance),
    augmented = map(covid_kclust, augment, covid_spm_location_km)
  )
```

```{r}
clusters <- 
  covid_kclusts %>%
  unnest(cols = c(tidied))

assignments <- 
  covid_kclusts %>% 
  unnest(cols = c(augmented))

clusterings <- 
  covid_kclusts %>%
  unnest(cols = c(glanced))
```

```{r}
clusterings %>%
  ggplot(aes(x = k, y = tot.withinss)) + 
  geom_line() + 
  geom_point() + ylab("") +
  ggtitle("Total Within Sum of Squares")
```

## Functions

#### `filter_continents` : filter covid dataset with `continent` variable for only sevon seperate continents 

```{r}
filter_continents <- function(covid_data) {
   covid_data <- covid_data %>%
     filter(continent == "Asia" |
           continent == "Africa" |
           continent == "Europe" |
           continent == "North America" |
           continent == "Oceania" |
           continent == "South America" |
           continent == "Antarctica"
           )
}

replace_all_na <- function(covid_data) {
  covid_data %>%
    replace(is.na(.), 0)
}
```

## EDA - Visualizations

#### New cases Smooth Per Million Over Time By Continent

```{r}
covid_spm %>%
  filter_continents %>%
  filter_world %>%
  filter(continent == "Asia") %>%
  ggplot(aes(x = date, y = new_cases_smoothed_per_million)) +
  geom_line() +
  geom_line(aes(color = location), show.legend = FALSE)
```

## Misc/EXperiments

```{r}
table(covid$location)
```

```{r}
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
data(penguins)

penguins_km <- penguins %>%
  drop_na(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) %>%
  select(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) %>%
  mutate(across(bill_length_mm:body_mass_g, scale))

set.seed(47)
penguin_kclust <- penguins_km %>%
  kmeans(centers = 3)

penguin_kclust
```

----------------------
