---
title: "COVID-EDA-Guy"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, cache=TRUE, fig.align = "center")
options(digits=5, pillar.sigfig = 7)
library(tidyverse)
library(tidymodels)
library(dplyr)
library(gganimate)
```

## Import Longitude-Latitude Dataset

```{r}
lats_long <- read.csv("https://raw.githubusercontent.com/albertyw/avenews/master/old/data/average-latitude-longitude-countries.csv")
```

```{r}
lats_long <- lats_long %>%
  rename(location = Country)
```

## Import John Hopkins Covid Dataset from 12/5/2021 named `covid`

```{r}
covid <- read.csv("https://raw.githubusercontent.com/ST47S-CompStats-Fall2021/GroupJ-COVID/main/data/covid-12-5-2021?token=AST7DELZDJCUNN4DQ752EDLBXAASG")

# change date to date variable
covid <- covid %>%
  mutate(date = as.Date(date))
```

## Functions

#### `filter_continents` : filter covid dataset with `continent` variable for only seven seperate continents 

```{r}
filter_continents <- function(covid_data) {
   covid_data <- covid_data %>%
     filter(continent == "Asia" |
           continent == "Africa" |
           continent == "Europe" |
           continent == "North America" |
           continent == "Oceania" |
           continent == "South America" |
           continent == "Antarctica"
           )
}
```

#### `only_countries` : filter covid dataset with `continent` variable for only sevon seperate continents 


```{r}
filter_world <- function(covid_data) {
  covid_data <- covid_data %>%
    filter(location != "World" |
           location != "Asia" |
           location != "Africa" |
           location != "Europe" |
           location != "North America" |
           location != "Oceania" |
           location != "South America" |
           location != "Antarctica"
           )
}
```

#### `replace_all_na` : replaces all `NA`'s in a dataframe with 0s

```{r}
replace_all_na <- function(covid_data) {
  covid_data %>%
    replace(is.na(.), 0)
}
```

```{r}
replace_0s_w1s <- function(covid_data) {
  covid_data %>%
    replace(.==(0), 1)
}
```

## Creating Sub-datasets

#### `covid_latest` : Only data from latest date (2020-)

```{r}
covid_latest <- covid %>%
  filter(date == max(date))
```

#### `covid_numeric` : Only numeric variables (and iso_code, location, date, continent)

```{r}
#select only numeric variables (outside of iso_code, location, and date)
covid_numeric <- cbind(
  covid %>%
    select(iso_code, location, date, continent),
  covid %>%
    select_if(is.numeric)
)
```

### `covid_per` : Only variables that containing `per` (and iso_code, location, date, continent)

```{r}
covid_per <- cbind(
  covid %>%
    select(iso_code, location, date, continent, population),
  covid %>%
    select(contains("per"))
)
```

### `covid_pm` : Only variables that end in per_million (and iso_code, location, date, continent)

```{r}
covid_pm <- cbind(
  covid %>%
    select(iso_code, location, date, continent),
  covid %>%
    select(ends_with("per_million"))
) %>%
  replace_all_na()
```

### `covid_spm` : Only variables that end with smoothed_per_million (and iso_code, location, date)

```{r}
covid_spm <- cbind(
  covid %>%
    select(iso_code, location, date, continent),
  covid %>%
    select(ends_with("smoothed_per_million"))
)
```

#### `covid_spm_location` : (pivot wider)

```{r}
covid_spm_location <- covid_spm %>%
  pivot_wider(names_from = date, values_from = !c(iso_code, date, location, continent)) %>%
  replace_all_na()
```

## EDA - Visualizations

#### New cases Per Million Over Time By Continent

```{r}
covid_pm %>%
  filter_continents %>%
  filter_world %>%
  ggplot(aes(x = date, y = new_cases_per_million)) +
  geom_line(aes(color = location), show.legend = FALSE) + 
  facet_wrap(~continent) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

```{r}
covid_pm %>%
  filter_continents %>%
  filter_world %>%
  ggplot(aes(x = date, y = new_cases_per_million)) +
  geom_line(aes(color = continent), show.legend = FALSE) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

#### Approximate excess mortality per million over time throughout covid

```{r}
covid_pm %>%
  filter_continents %>%
  ggplot(aes(x = date, y = excess_mortality_cumulative_per_million)) +
  geom_smooth(aes(color = continent), se = FALSE)
```

#### HDI-covid plots

```{r}
# data

HDI_2020 <- read.csv("https://raw.githubusercontent.com/ST47S-CompStats-Fall2021/GroupJ-COVID/main/data/World_Bank_2020_HDI.csv?token=AST7DELBTYIJUK2GTQ7TXNDBXOZWA")

HDI_2020 <- HDI_2020 %>%
  rename(location = ï..Country.Name) %>%
  rename(HDI = X2020)

covid_latest_hdi <- left_join(covid_latest, HDI_2020, by = "location")

covid_latest_hdi <- covid_latest_hdi %>%
  filter(!is.na(HDI)) %>%
  arrange(desc(HDI))

covid_pm_hdi <- left_join(covid_pm, HDI_2020, by = "location") %>%
  filter(!is.na(HDI))
```

```{r}
covid_latest_hdi %>%
  ggplot(aes(x = HDI, y = total_cases_per_million)) +
  geom_point(aes(color = continent)) +
  geom_smooth(se = FALSE)
```

```{r}
covid_latest_hdi %>%
  ggplot(aes(x = HDI, y = total_cases_per_million)) +
  geom_point(aes(color = continent)) +
  facet_wrap(~continent) +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
library(gapminder)
```

```{r}
hdi_animate <- ggplot(covid_pm_hdi, aes(HDI, total_cases_per_million, color = location)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_color_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  facet_wrap(~continent) +
  labs(title = 'Date: {frame_time}', x = 'HDI', y = 'total_cases_per_million') +
  transition_time(date) +
  ease_aes('linear') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

animate(hdi_animate, duration = 10)
```

#### Tourism-covid plots

```{r}
# data
inter_arrivals <- read.csv("https://raw.githubusercontent.com/ST47S-CompStats-Fall2021/GroupJ-COVID/main/data/World_Bank_International_Arrival.csv?token=AST7DEIUKOA6UO5AO6ETDVDBXO4M2")

IA_2019 <- inter_arrivals %>%
  rename(location = ï..Country.Name) %>%
  rename(inter_arrivals = X2019) %>%
  select(location, inter_arrivals)

covid_pm_IA <- left_join(covid_pm, IA_2019, by = "location")

covid_pm_IA <- covid_pm_IA %>%
  filter(date == max(date)) %>%
  filter(!is.na(inter_arrivals)) %>%
  arrange(desc(inter_arrivals))
```

```{r}
covid_pm_IA %>%
  ggplot(aes(x = inter_arrivals, y = total_cases_per_million)) +
  geom_point(aes(color = continent)) +
  geom_smooth(se = FALSE)
```

```{r}
covid_pm_IA %>%
  ggplot(aes(x = inter_arrivals, y = total_cases_per_million)) +
  geom_point(aes(color = continent)) +
  facet_wrap(~continent, scales = "free") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

```{r}
covid_pm_IA %>%
  filter(continent == "Africa") %>%
  ggplot(aes(x = inter_arrivals, y = total_cases_per_million)) +
  geom_point(aes(color = location)) +
  facet_wrap(~continent, scales = "free") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

#### GDP

```{r}
gdp <- read.csv("https://raw.githubusercontent.com/ST47S-CompStats-Fall2021/GroupJ-COVID/main/data/World_Bank_gdp_per_capita.csv?token=AST7DEMK7QPXWCAXKX35FXDBXPJOG")

gdp_2019 <- gdp %>%
  rename(location = ï..Country.Name) %>%
  rename(gdp = X2019) %>%
  select(location, gdp)

covid_per_gdp <- left_join(covid_per, gdp_2019, by = "location")

covid_per_gdp_latest <- covid_per_gdp %>%
  filter(date == max(date)) %>%
  filter(!is.na(gdp))
```

```{r}
covid_per_gdp_latest %>%
  filter(!is.na(total_vaccinations_per_hundred)) %>%
  filter_continents() %>%
  filter_world() %>%
  mutate(gdp = log(gdp)) %>%
  mutate(vax = if_else(total_vaccinations_per_hundred == 0, "no", "yes")) %>%
  ggplot(aes(x = gdp, y = total_vaccinations_per_hundred)) +
  geom_point(aes(color = continent)) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_text(aes(label=location),hjust=0, vjust=0, size = 3) +
  xlab("log(gdp per capita)")
```

## Misc/Experiments

----------------------









