---
title: "pca, clustering, maps, and viz"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidymodels)
library(lubridate)
library(GGally)
```

```{r}
filter_world <- function(covid_data) {
  covid_data <- covid_data %>%
    filter(location != "World" |
           location != "Asia" |
           location != "Africa" |
           location != "Europe" |
           location != "North America" |
           location != "Oceania" |
           location != "South America" |
           location != "Antarctica"
           )
}
filter_continents <- function(covid_data) {
   covid_data <- covid_data %>%
     filter(continent == "Asia" |
           continent == "Africa" |
           continent == "Europe" |
           continent == "North America" |
           continent == "Oceania" |
           continent == "South America" |
           continent == "Antarctica"
           )
}
```

```{r}
covid <- read.csv("https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/owid-covid-data.csv")

vars <- names(covid)

vars

vars <- vars[! vars %in% c('date')]

sum <- covid %>% group_by(date) %>% 
  summarise(across(vars, ~ sum(is.na(.))))

prop <- covid %>% group_by(date) %>% 
  summarise(across(vars, funs(sum(is.na(.)) / length(.))))

test <- covid %>% filter(date == "2020-01-01")

NAsummary <- prop %>% 
  pivot_longer(!date, names_to="variable") %>%
  mutate(date=ymd(date))

NAsummary %>% filter(grepl('case', variable)) %>% ggplot(aes(x=date, y=value, color=variable)) + geom_line()

NAsummary %>% ggplot(aes(x=date, y=value, color=variable)) + 
  geom_line() + 
  theme(legend.position = "none")

covid %>% filter(continent==c("Africa", "Asia", "Europe", "North America", "Oceania", "South America")) %>% ggplot(aes(x=date, y=new_cases, color = continent)) + 
  geom_point() 


one_day <- covid %>% filter(date == "2021-12-05")

data.frame(colSums(is.na(one_day)))

```

```{r}
library(tidytext)

replace_all_na <- function(covid_data) {
  covid_data %>%
    replace(is.na(.), 0)
}

one_day_prepped <- replace_all_na(one_day) %>%
  filter_continents() %>%
  filter_world()

one_day_prepped %>% summarise_all(mean)

ready <- one_day_prepped %>% select(!c(excess_mortality_cumulative, excess_mortality, excess_mortality_cumulative_per_million, excess_mortality_cumulative_absolute, reproduction_rate))

pca_recipe <- recipe(~., data = ready) %>%
  step_center(all_numeric()) %>%
  step_scale(all_numeric()) %>%
  step_pca(all_numeric(), id = "pca")

pca_recipe

pca_estimates <- prep(pca_recipe)

juice(pca_estimates)

juice(pca_estimates) %>%
  ggplot(aes(PC3, PC5)) +
  geom_point(aes(color = continent), alpha = 0.7, size = 2)+
  labs(title="PCA from tidymodels") + 
  xlim(-5, 5) + ylim(-5, 5) +
  geom_smooth(method = "lm", se = FALSE)

tidied_pca <- tidy(pca_estimates, 2)

tidy_pca_loadings <- pca_estimates%>% 
  tidy(id = "pca")

tidy_pca_loadings

tidy_pca_loadings %>%
  filter(component %in% paste0("PC", 1:5)) %>%
  mutate(component = fct_inorder(component)) %>%
  ggplot(aes(value, terms, fill = terms)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~component, nrow = 1) +
  labs(y = NULL)

plot_loadings <- tidy_pca_loadings %>% 
  filter(component %in% c("PC1")) %>% 
  mutate(terms = tidytext::reorder_within(terms, 
                                          abs(value), 
                                          component)) %>% 
  ggplot(aes(abs(value), terms, fill = value>0)) +
  geom_col() +
  facet_wrap( ~component, scales = "free_y") +
  scale_y_reordered() + # appends ___ and then the facet at the end of each string
  scale_fill_manual(values = c("deepskyblue4", "darkorange")) +
  labs( x = "absolute value of contribution",
        y = NULL,
        fill = "Positive?",
        title = "PCA Loadings Plot") +
  theme_minimal()


plot_loadings
```

```{r}
juice_df <- juice(pca_estimates)

head(juice_df)
```


```{r}
juice_df %>%
  ggplot(aes(PC1,PC2)) +
  geom_point(aes(color = continent), alpha = 0.7, size = 2)+
  labs(title="PCA from tidymodels") + 
  xlim(-5, 5) + ylim(-5, 5) +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
juice_df %>%
  ggplot(aes(PC1,PC3)) +
  geom_point(aes(color = continent), alpha = 0.7, size = 2)+
  labs(title="PCA from tidymodels") + 
  xlim(-5, 5) + ylim(-5, 5) +
  geom_smooth(method = "lm", se = FALSE)
```
```{r}
juice_df %>%
  select(starts_with("PC"), continent) %>%
  ggpairs(aes(color = continent))
```
As we can see, there is no immediately obvious groupings between the continents based on the principle componenents.

```{r}
set.seed(47)
pca_kclust <- juice_df %>%
  select(starts_with("PC")) %>%
  kmeans(centers=6)

pca_kclust
```
```{r}
loc_clusters <- pca_kclust %>% augment(juice_df)
loc_clusters
```
```{r}
# getting long-lat data
lats_long <- read.csv("https://raw.githubusercontent.com/albertyw/avenews/master/old/data/average-latitude-longitude-countries.csv")

lats_long <- lats_long %>%
  rename(location = Country)
```

combining long-lat data
```{r}
long_lat_clusters <- left_join(loc_clusters, lats_long, by="location")
head(long_lat_clusters)
```
Now, we plot everything on a map, this time coloring the locations based on their cluster
```{r}
world <- map_data("world")

map <- ggplot() +
  geom_map(
    data = world, map = world,
    aes(long, lat, map_id = region),
    color = "black", fill = "lightgray", size = 0.01
  ) +
  geom_point(data = long_lat_clusters, aes(x=Longitude, y=Latitude, color=.cluster))

map
```


## PCA-Clustering by Day

vaccinations started 2020-12-16

We want to loop through the days of the data since 2020-12-16. To do this, we first convert the `date` variable from a factor to a type `Date`, and increment for each loop. We maintain some global dataframe that consists of the dates we have already iterated through. For each iteration, we call rbind with the global dataframe and the local dataframe that contains only information from a specific data. To create the local dataframe, we will be performing PCA on the data for a given day, and then we will perform clustering based on the PCAs and augment the cluster labels with the non-PCA. After iterating through all the days, we will have a dataframe with columns `Location`, `date`, `cluster`. At this point, we will cbind this data with the longitude and latitude data. The resultant dataframe will then allow us to plot on a map of the world the cluster labels, so that we can try to gain a sense of whether there is a spatial relationship between clusters. We will create an animation that shows the clusters based on the PCAs over time. Our hypothesis is that data points in the same location will tend to be in the same cluster, even as time progresses. Furthermore, we predict that changes in groupings will also occur based on spatial relations. Note that clusters will have no relation at all between timesteps. Therefore, we will not measure whether or not the color of a certain location is different between frames, but whether the (arbitrary) color of a point is the same as the color of points near it even as time progresses.

### converting date data into date type
```{r}
cleaned_covid <- covid %>%
  replace_all_na() %>%
  filter_continents() %>%
  filter_world() %>%
  mutate(date = as.Date(date))
```

```{r}
cleaned_covid %>% head()
```
Need to check dates such that the range for none of the variables is 
```{r}
test_cleaned <- cleaned_covid %>% 
  select(!is.character) %>%
  select(!c(excess_mortality_cumulative, excess_mortality, excess_mortality_cumulative_per_million, excess_mortality_cumulative_absolute, reproduction_rate, total_boosters, total_boosters_per_hundred, weekly_icu_admissions, weekly_icu_admissions_per_million, weekly_hosp_admissions, weekly_hosp_admissions_per_million)) %>%
  filter(date > as.Date("2020-12-28")) %>%
  group_by(date) %>%
  summarize_all(sd) %>%
  filter_all(all_vars(. != 0))

head(test_cleaned)
```

Above, we can see the bad variables that we need to exclude for our analysis. We save this cleaner dataset:

```{r}
cleaner_covid <- cleaned_covid %>%
  select(!c(excess_mortality_cumulative, excess_mortality, excess_mortality_cumulative_per_million, excess_mortality_cumulative_absolute, reproduction_rate, total_boosters, total_boosters_per_hundred, weekly_icu_admissions, weekly_icu_admissions_per_million, weekly_hosp_admissions, weekly_hosp_admissions_per_million, tests_units)) 

cleanest_covid <- cleaner_covid %>%
  filter(date > as.Date("2020-12-28"))
```

Note that 2020-12-28 may be too late a date. We may want to view the data even before this.
```{r}
head(cleaner_covid)
```

```{r}
sorted_dates <- sort(unique(test_cleaned$date))
length(sorted_dates)
```

```{r}
first_day_df <- cleaner_covid %>% filter(date == as.Date("2020-12-28"))
first_day_df %>%
  group_by(date) %>%
  summarize_all(sd)
```
(Yes, I know this is messy, inefficient code. It only gets worse from here)

```{r}
# function to perform PCA and cluster based
get_cluster_df <- function(day_df) {
  pca_recipe <- recipe(~., data=day_df) %>%
    step_center(all_numeric()) %>%
    step_scale(all_numeric()) %>%
    step_pca(all_numeric(), id="pca")
  
  pca_estimates <- prep(pca_recipe)
  juice_df <- juice(pca_estimates)
  
  pca_kclust <- juice_df %>%
    select(starts_with("PC")) %>%
    kmeans(centers=6)
  
  loc_clusters <- pca_kclust %>% 
    augment(juice_df)
  
  return(loc_clusters)
}
```

```{r}
set.seed(4700)
# the first date
all_loc_clusters <- get_cluster_df(first_day_df)
head(all_loc_clusters)
```
### CHECKPOINT

```{r}
set.seed(4747)
  
# the rest of the dates
for (day in sorted_dates){
  day_data <- cleaner_covid %>%
    filter(date == day)
  
  loc_clusters <- get_cluster_df(day_data)
  all_loc_clusters <- rbind(all_loc_clusters, loc_clusters)
}
```

lets gooooooooo

```{r}
head(all_loc_clusters)
```

```{r}
all_long_lat_clusters <- left_join(all_loc_clusters, lats_long, by="location")
head(all_long_lat_clusters)
```


```{r, eval=FALSE}
library(gapminder)
library(gganimate)

map_anim <- ggplot() +
  geom_map(
    data = world, map = world,
    aes(long, lat, map_id = region),
    color = "black", fill = "lightgray", size = 0.01
  ) +
  geom_point(data = all_long_lat_clusters, aes(x=Longitude, y=Latitude, color=.cluster)) +
  transition_time(date)

animate(map_anim, duration = 20)
```
```{r, eval=FALSE}
anim_save("/Users/jaredmejia/Documents/Pomona/fall21/compstats/GroupJ-COVID/workbooks/pca-cluster-map.gif")
```

TODO:
Okay, clearly we need to do something to make sure the colors are consistent throughout. One idea: 
  Identify certain countries that are always in the "same" cluster, such as the United States, Central African Republic, and Germany. Then, arbitrarily swap the names of the columns such that the cluster that contains the US is always labeled PC1, the cluster that contains the Central African Republic is always labeled PC2, and the cluster that contains Germany is always labeled PC3.

Also, consider changing cluster count to 5 from 6.

```{r}
test_day_cluster <- get_cluster_df(first_day_df)
head(test_day_cluster)
```
```{r}
test_long_lat_clusters <- left_join(test_day_cluster, lats_long, by="location")
head(test_long_lat_clusters)

world <- map_data("world")

map1 <- ggplot() +
  geom_map(
    data = world, map = world,
    aes(long, lat, map_id = region),
    color = "black", fill = "lightgray", size = 0.01
  ) +
  geom_point(data = test_long_lat_clusters, aes(x=Longitude, y=Latitude, color=.cluster))

map1
```

We can create a new column that sets the color of all clusters with the US as red, sets all clusters with CAR as blue, sets all clusters with Germany as green, sets the three other clusters as purple and pink and yellow. 
```{r}
a <- test_day_cluster %>%
  filter(location == "United States")

a
```
```{r}
us_clust = a$.cluster
us_clust
```


```{r}
test_long_lat_clusters <- test_long_lat_clusters %>%
  mutate(color = ifelse(.cluster == us_clust, "red", "orange1"))

map <- ggplot() +
  geom_map(
    data = world, map = world,
    aes(long, lat, map_id = region),
    color = "black", fill = "lightgray", size = 0.01
  ) +
  geom_point(data = test_long_lat_clusters, aes(x=Longitude, y=Latitude), color=test_long_lat_clusters$color)

map
```

```{r}

us_row <- test_day_cluster %>%
  dplyr::filter(location == "United States")
us_clust = us_row$.cluster

car_row <- test_day_cluster %>%
  dplyr::filter(location == "Central African Republic")
car_clust = car_row$.cluster

ger_row <- test_day_cluster %>%
  dplyr::filter(location == "Germany") 
ger_clust = ger_row$.cluster

remaining_clusts <- setdiff(as.factor(seq(1,6)), c(us_clust, car_clust, ger_clust))

test_long_lat_clusters <- test_long_lat_clusters %>%
  mutate(color = ifelse(.cluster == us_clust, "red", "orange1")) %>%
  mutate(color = ifelse(.cluster == car_clust, "steelblue2", color)) %>%
  mutate(color =  ifelse(.cluster == ger_clust, "orchid2", color)) %>%
  mutate(color = ifelse(.cluster == remaining_clusts[1], "springgreen3", color)) %>%
  mutate(color = ifelse(.cluster == remaining_clusts[2], "purple3", color))

map <- ggplot() +
  geom_map(
    data = world, map = world,
    aes(long, lat, map_id = region),
    color = "black", fill = "lightgray", size = 0.01
  ) +
  geom_point(data = test_long_lat_clusters, aes(x=Longitude, y=Latitude), color=test_long_lat_clusters$color, show.legend = TRUE)

map
```
```{r}
map1
```
```{r}
test_day_cluster %>%
  select(iso_code) %>%
  dplyr::filter(iso_code == "ALB")
```

New function with color setting
```{r}
# function to perform PCA and cluster based and colors!
get_cluster_colored_df <- function(day_df) {
  pca_recipe <- recipe(~., data=day_df) %>%
    step_center(all_numeric()) %>%
    step_scale(all_numeric()) %>%
    step_pca(all_numeric(), id="pca")
  
  pca_estimates <- prep(pca_recipe)
  juice_df <- juice(pca_estimates)
  
  pca_kclust <- juice_df %>%
    select(starts_with("PC")) %>%
    kmeans(centers=6)
  
  loc_clusters <- pca_kclust %>% 
    augment(juice_df)
  
  us_row <- loc_clusters %>%
    dplyr::filter(location == "United States")
  us_clust = us_row$.cluster
  
  car_row <- loc_clusters %>%
    dplyr::filter(location == "Central African Republic")
  car_clust = car_row$.cluster
  
  ger_row <- loc_clusters %>%
    dplyr::filter(location == "Germany") 
  ger_clust = ger_row$.cluster
  
  ;costa_row <- loc_clusters %>%
    dplyr::filter(location == "Costa Rica")
  costa_clust = costa_row$.cluster
  
  remaining_clusts <- setdiff(as.factor(seq(1,6)), c(us_clust, car_clust, ger_clust, costa_clust))
  
  loc_clusters <- loc_clusters %>%
    mutate(color = ifelse(.cluster == us_clust, "red", "orange1")) %>%
    mutate(color = ifelse(.cluster == costa_clust, "springgreen3", color)) %>%
    mutate(color = ifelse(.cluster == remaining_clusts[1], "purple3", color)) %>%
        mutate(color =  ifelse(.cluster == ger_clust, "orchid2", color)) %>%
        mutate(color = ifelse(.cluster == car_clust, "steelblue2", color))
  
  return(loc_clusters)
}
```

```{r}
set.seed(4700)
# the first date
all_loc_clusters_colored <- get_cluster_colored_df(first_day_df)
head(all_loc_clusters_colored)
```
### CHECKPOINT

```{r}
set.seed(4747)
  
# the rest of the dates
for (day in sorted_dates){
  day_data <- cleaner_covid %>%
    dplyr::filter(date == day)
  
  loc_clusters <- get_cluster_colored_df(day_data)
  all_loc_clusters_colored <- rbind(all_loc_clusters_colored, loc_clusters)
}
```

```{r}
all_loc_clusters_colored %>%
  dplyr::filter(color != "red", color != "steelblue2", continent=="Europe") %>%
  dplyr::count(location) 
```



```{r}
all_long_lat_clusters_colored <- left_join(all_loc_clusters_colored, lats_long, by="location")
head(all_long_lat_clusters_colored)
```

```{r, eval=FALSE}
map_anim_colored <- ggplot() +
  geom_map(
    data = world, map = world,
    aes(long, lat, map_id = region),
    color = "black", fill = "lightgray", size = 0.01
  ) +
  geom_point(data = all_long_lat_clusters_colored, aes(x=Longitude, y=Latitude), color=all_long_lat_clusters_colored$color) +
  transition_time(date) +
  labs(title = 'PCA Clusters on Date: {frame_time}', x = '', y = '')

animate(map_anim_colored, duration = 20)
```

```{r, eval=FALSE}
anim_save("/Users/jaredmejia/Documents/Pomona/fall21/compstats/GroupJ-COVID/workbooks/pca-cluster-map-colored.gif")
```



Now, we shall repeat this process, but getting rid of all the variables relating to population size

```{r}
cleaner_covid_2 <- cleaned_covid %>%
  select(contains("per"), c(continent, location, date, aged_65_older, aged_70_older, gdp_per_capita, extreme_poverty, cardiovasc_death_rate, diabetes_prevalence, female_smokers, male_smokers, handwashing_facilities, hospital_beds_per_thousand, life_expectancy, human_development_index)) %>%
  select(!c(excess_mortality_cumulative_per_million, total_boosters_per_hundred, weekly_icu_admissions_per_million, weekly_hosp_admissions_per_million))

cleanest_covid_2 <- cleaner_covid_2 %>%
  dplyr::filter(date > as.Date("2020-12-28")) %>%
  dplyr::filter(date < as.Date("2021-12-10"))
```

```{r}
head(cleaner_covid_2)
```
### CHECKPOINT

```{r}
# function to perform PCA and cluster based and colors!
get_cluster_colored_df2 <- function(day_df) {
  pca_recipe <- recipe(~., data=day_df) %>%
    step_center(all_numeric()) %>%
    step_scale(all_numeric()) %>%
    step_pca(all_numeric(), id="pca")
  
  pca_estimates <- prep(pca_recipe)
  juice_df <- juice(pca_estimates)
  
  pca_kclust <- juice_df %>%
    select(starts_with("PC")) %>%
    kmeans(centers=6)
  
  loc_clusters <- pca_kclust %>% 
    augment(juice_df)
  
  tunisia_row <- loc_clusters %>%
    dplyr::filter(location == "Tunisia")
  tunisia_clust = tunisia_row$.cluster
  
  car_row <- loc_clusters %>%
    dplyr::filter(location == "Central African Republic")
  car_clust = car_row$.cluster
  
  ger_row <- loc_clusters %>%
    dplyr::filter(location == "Germany") 
  ger_clust = ger_row$.cluster
  
  uae_row <- loc_clusters %>%
    dplyr::filter(location == "United Arab Emirates")
  uae_clust = uae_row$.cluster
  
  remaining_clusts <- setdiff(as.factor(seq(1,6)), c(tunisia_clust, car_clust, ger_clust, uae_clust))
  
  loc_clusters <- loc_clusters %>%
    mutate(color = ifelse(.cluster == remaining_clusts[1], "red", "orange1")) %>%
    mutate(color = ifelse(.cluster == uae_clust, "purple3", color)) %>%
        mutate(color = ifelse(.cluster == tunisia_clust, "springgreen3", color)) %>%
        mutate(color =  ifelse(.cluster == ger_clust, "orchid2", color)) %>%
        mutate(color = ifelse(.cluster == car_clust, "steelblue2", color))
  
  return(loc_clusters)
}
```

```{r}
set.seed(4700)
# the first date
first_day_df2 <- cleaner_covid_2 %>% dplyr::filter(date == as.Date("2020-12-28"))
all_loc_clusters_colored2 <- get_cluster_colored_df2(first_day_df2)
head(all_loc_clusters_colored2)
```

```{r}
set.seed(4747)
  
# the rest of the dates
for (day in sorted_dates){
  day_data <- cleaner_covid_2 %>%
    dplyr::filter(date == day)
  
  loc_clusters <- get_cluster_colored_df2(day_data)
  all_loc_clusters_colored2 <- rbind(all_loc_clusters_colored2, loc_clusters)
}
```

```{r}
all_loc_clusters_colored2 %>%
  dplyr::filter(color!="red", color != "orchid2", color != "steelblue2", color != "purple3") %>%
  dplyr::count(location) %>%
  arrange(desc(n))
```

```{r}
all_long_lat_clusters_colored2 <- left_join(all_loc_clusters_colored2, lats_long, by="location")
head(all_long_lat_clusters_colored2)
```

```{r, eval=FALSE}
library(gganimate)
library(gapminder)

map_anim_colored2 <- ggplot() +
  geom_map(
    data = world, map = world,
    aes(long, lat, map_id = region),
    color = "black", fill = "lightgray", size = 0.01
  ) +
  geom_point(data = all_long_lat_clusters_colored2, aes(x=Longitude, y=Latitude), color=all_long_lat_clusters_colored2$color) +
  transition_time(date) +
  labs(title = 'PCA Clusters on Date: {frame_time}', x = '', y = '')

animate(map_anim_colored2, duration = 20)
```

```{r, eval=FALSE}
anim_save("/Users/jaredmejia/Documents/Pomona/fall21/compstats/GroupJ-COVID/workbooks/pca-cluster-map-colored-nopop.gif")
```














